---
title: "Velocity computation"
output: rmarkdown::html_vignette
bibliography: methods.bib
vignette: >
  %\VignetteIndexEntry{Velocity computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(saccadr)
```

Because the gaze samples tend to be noisy, different methods use various approaches for computing velocity from noisy samples. Methods by @EngbertKliegl2003 and @Otero-Millan2014 used the same approach based on averaging over multiple samples to compute velocity, whereas @NystromHolmqvist2010 compute a simple derivative and then filter it. By default, package uses the former approach (`velocity_function = diff_ek`) but you can also use the latter (`velocity_function = diff_nh`) or implement a custom method (see below). Acceleration is computed the same way but from velocity samples.

## Differentiation following @EngbertKliegl2003
This method computes velocity over a predefined time window controlled via `ek_velocity_time_window` that defaults to 20 ms. The width of the window is converted to samples based on the `sampling_rate` parameter and is ensured to be an _odd_ number equal to or larger than 3. I.e., for a window that is three samples long, the velocity is computed based on one sample before and one sample after the current, for a five-samples window - two samples before and after, etc. The code always starts with the requested window size but all iteratively reduces it by two samples to accommodate for trial limits and missing values. An `NA` velocity is returned only if even the smallest window (three samples wide) cannot be used.

In general, horizontal and vertical velocity components are computed as
$$v_x[i] = \frac{\sum_{j=1}^{(N-1)/2}x[i+j] - x[i-j]}{\sum_{j=1}^{(N-1)/2}2j\cdot\Delta t}$$
where $i$ is the index of a sample, $\Delta t = \frac{1}{sample~rate}$ is a duration of a single sampling frame, and $N$ is an _odd_ integer width of the moving average used to compute the velocity. In @EngbertKliegl2003, $N=5$ and $\Delta t = 4$ (250 Hz sampling rate) that translates into a 20 ms moving average window (default value used in the method implementation). Below is a derivation that shows the equivalence of the formula above to formula 1 in @EngbertKliegl2003.  For $N=5$:
$$\frac{\sum_{j=1}^{(5-1)/2}x[i+j] - x[i-j]}{\sum_{j=1}^{(5-1)/2}2j\cdot\Delta t}=$$
$$\frac{\sum_{j=1}^{2}x[i+j] - x[i-j]}{\sum_{j=1}^{2}2j\cdot\Delta t}=$$

$$\frac{x[i+1] - x[i-1]+x[i+2] - x[i-2]}{2\Delta t + 4\Delta t}=$$
$$\frac{x[i+2] + x[i+1] - x[i-1] - x[i-2]}{6\Delta t}$$

## Differentiation following @NystromHolmqvist2010
Horizontal and vertical components of velocity are computed from adjacent samples, the amplitude is computed from the components and then is filtered using [Savitzky-Golay filter](https://en.wikipedia.org/wiki/Savitzky%E2%80%93Golay_filter). Filter order is controlled via `nh_sg_filter_order` option and defaults to `2`.

## Comparing two methods
The two methods produce comparable but different velocity values. However, this has only a minor effect on saccade detection, see vignette on _Using sample votes_.

```{r warning=FALSE, fig.fullwidth=TRUE, out.width="100%", fig.dim = c(8, 4)}
# computing velocity using two methods
vel_ek <- saccadr::diff_ek(x = single_trial$x,
                           y = single_trial$y,
                           trial = rep(1, nrow(single_trial)),
                           sample_rate = 500)
vel_nh <- saccadr::diff_nh(x = single_trial$x,
                           y = single_trial$y,
                           trial = rep(1, nrow(single_trial)),
                           sample_rate = 500)
vel_comparison <- na.omit(data.frame(EK = vel_ek[['amp']], NH = vel_nh[['amp']])) %>%
  dplyr::filter(NH > 0) %>% # values can be negative due to filtering
  dplyr::mutate(logEK = log(EK), logNH = log(NH)) 

# plotting amplitude comparison
pearson_rho <- cor(vel_comparison[['EK']], vel_comparison[['NH']])

ggplot(data = vel_comparison, aes(x = EK, y = NH)) +
  geom_abline() +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(subtitle = sprintf("Pearson's ρ = %.2f\nPearson's ρ for log transformed velocities = %.2f", pearson_rho, cor(vel_comparison[['logEK']], vel_comparison[['logNH']])),
       x = "Vek",
       y = "Vnh")

# plotting 
ggplot(data = NULL, aes(x = vel_comparison[['EK']] - vel_comparison[['NH']])) +
  geom_histogram(bins = 50) +
  xlab("Vek - Vnh") +
  labs(title = "Difference in velocities computed via two methods")
```

## Implementing and using a custom differentiation function
A custom function should be as a follows

```{r eval = FALSE}
diff_custom <- function(x, y,  trial, sample_rate, options=NULL) {
  # compute vertical and horizontal components
  df <- data.frame(x = ..., # differentiate values of x 
                   y = ...) # differentiate values of y
  
  # compute amplitude
  df[['amp']] = sqrt(df[['x']]^2 + df[['y']]^2))

  # return a data.frame with columns x, y, and amp
  return(df)
}
```

Arguments that are passed to the function (all vectors have the same length):

* `x`, `y` vectors with values, either x and y gaze coordinates or velocity components (for acceleration).
* `sample_rate` scalar value in Hz.
* `trial` vector with trial index per sample.
* `options` a named list with method specific options. See example below for how to use them. See also `option_or_default()` function.

The function must return a data.frame with three columns: `x` (horizontal component), `y` (vertical component), and `amp` (amplitude). 

Here is an example of a simple velocity function that respects trial borders
```{r eval = FALSE}
library(dplyr)

diff_custom <- function(x, y,  trial, sample_rate, options=NULL) {
  # computing frame time step 
  delta_t_s <- 1 / sample_rate
  
  # --- differentiate (compute velocity or acceleration) and filter
  data.frame(trial = trial,
             x = x,
             y = y) %>%

    # compute velocity and acceleration for each trial
    group_by(trial) %>%
    mutate(x = (x - lag(x)) / delta_t_s,
           y = (y - lag(y)) / delta_t_s,
           amp = sqrt(x^2 + y^2))
}
```

And here is a similar function but that uses an optional `custom_velocity_lag` parameter, so that it can be different from 1
```{r eval = FALSE}
diff_lag_n <- function(x, y,  trial, sample_rate, options=NULL) {
  # obtain lag or use default
  lag_n <- saccadr::option_or_default(options, "custom_velocity_lag", 1)
  
  # computing frame time step 
  delta_t_s <- 1 / sample_rate
  
  # --- differentiate (compute velocity or acceleration) and filter
  data.frame(trial = trial,
             x = x,
             y = y) %>%

    # compute velocity and acceleration for each trial
    group_by(trial) %>%
    mutate(x = (x - lag(x, lag_n)) / delta_t_s,
           y = (y - lag(y, lag_n)) / delta_t_s,
           amp = sqrt(x^2 + y^2))
}
```

## References